---
title: "Taxonomy Report"
author: "López-Chablé Edgard Uriel"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: zenburn
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message = FALSE, warning = FALSE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(stringr)
library(plotly)
library(networkD3)
library(htmlwidgets)
library(webshot)
library(magick)
```

## 1. Data Import and Preprocessing

We import the Kraken2 taxonomy report and perform initial cleanup, such as removing leading/trailing whitespaces from the taxon names.

```{r data_import}
# Create output directory for high-res figures
out_dir <- "../results/taxonomy"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Read Kraken2 report
kraken_data <- read.delim("../data/taxonomy/reporte_taxonomia.txt",
                          header = FALSE, 
                          sep = "\t", 
                          col.names = c("Percentage", "Clade_Reads", "Taxon_Reads", 
                                        "Rank", "TaxID", "Name"))

# Clean up whitespace used by Kraken for indentation
kraken_data <- kraken_data %>%
  mutate(Name = str_trim(Name))

head(kraken_data)
```

---

## 2. Global Species Abundance

We filter the data strictly to the Species level (Rank 'S') to visualize the top 10 most abundant species in our sample.

```{r relative_abundance_bar}
# Filter and sort for the top 10 species
bar_data <- kraken_data %>%
  filter(Rank == "S") %>%
  arrange(desc(Percentage)) %>%
  head(10)

# Create the bar plot
p_bar <- ggplot(bar_data, aes(x = reorder(Name, Percentage), y = Percentage, fill = Name)) +
  geom_bar(stat = "identity", color = "black", show.legend = FALSE) +
  coord_flip() + 
  theme_minimal() +
  labs(title = "Relative Species Abundance in Control Sample",
       subtitle = "Taxonomic classification based on k-mers (Kraken2)",
       x = "Identified Species",
       y = "Percentage of Reads (%)") +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#2c3e50"),
    plot.subtitle = element_text(size = 12, color = "#7f8c8d"),
    axis.text.y = element_text(face = "italic", size = 12, color = "black")
  ) +
  scale_fill_viridis_d(option = "mako", direction = -1)

p_bar

# Export to PDF and 600 DPI TIFF
ggsave(filename = paste0(out_dir, "/barplot_taxonomy.pdf"), plot = p_bar, width = 10, height = 8)
ggsave(filename = paste0(out_dir, "/barplot_taxonomy.tiff"), plot = p_bar, width = 10, height = 8, dpi = 600, compression = "lzw")
```

---

## 3. Lowest Common Ancestor (LCA) Read Distribution

Not all reads assigned to the Genus level can be resolved to a specific Species. This Sankey diagram visualizes the flow of reads and highlights the proportion of sequences conserved at the Genus level due to the LCA algorithm.

```{r lca_sankey}
# Extract total reads for the Pectobacterium genus
genus_reads <- kraken_data %>% 
  filter(Rank == "G" & Name == "Pectobacterium") %>% 
  pull(Clade_Reads)

# Extract breakdown of specific Pectobacterium species
species_data <- kraken_data %>% 
  filter(Rank == "S" & grepl("Pectobacterium", Name)) %>%
  select(Name, Clade_Reads)

# Calculate remaining reads (LCA algorithm resolution limit)
sum_species <- sum(species_data$Clade_Reads)
unclassified_reads <- genus_reads - sum_species

# Build Sankey Links
links <- data.frame(
  source = rep("Pectobacterium (Total Genus)", nrow(species_data) + 1),
  target = c(species_data$Name, "Conserved at Genus Level (LCA)"),
  value = c(species_data$Clade_Reads, unclassified_reads)
)

# Build Nodes and assign numerical IDs required by networkD3
nodes <- data.frame(name = unique(c(links$source, links$target)))

links$IDsource <- match(links$source, nodes$name) - 1 
links$IDtarget <- match(links$target, nodes$name) - 1

# Draw interactive Sankey Network
sankey_plot <- sankeyNetwork(Links = links, Nodes = nodes,
                             Source = "IDsource", Target = "IDtarget",
                             Value = "value", NodeID = "name",
                             units = " Reads", 
                             fontSize = 14, nodeWidth = 30, 
                             sinksRight = FALSE, 
                             margin = list(left = 50, right = 50))

sankey_plot

# Export Interactive Sankey to Static High-Res Formats
# 1. Save as temporary HTML
saveWidget(sankey_plot, "temp_sankey.html", selfcontained = FALSE)

# 2. Capture as PDF
webshot("temp_sankey.html", file = paste0(out_dir, "/sankey_lca.pdf"))

# 3. Read PDF and render as 600 DPI TIFF using magick
img_sankey <- image_read_pdf(paste0(out_dir, "/sankey_lca.pdf"), density = 600)
image_write(img_sankey, paste0(out_dir, "/sankey_lca.tiff"), format = "tiff")

# Clean up temporary HTML files
unlink(c("temp_sunburst.html", "temp_sankey.html", "temp_sunburst_files", "temp_sankey_files"), recursive = TRUE)
```

---

## Session Information
```{r session_info}
sessionInfo()
```


