---
title: "Exploratory Analysis"
author: "López-Chablé Edgard Uriel"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: zenburn
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message = FALSE, warning = FALSE}
# Load necessary packages
library(cowplot)
library(viridis)
library(tximport)
library(DESeq2)
library(tidyverse)
library(ggrepel)

# TIFF output directory (created automatically)
tiff_dir <- "../results/TIFFs/1_ExploratoryAnalysis"
dir.create(tiff_dir, showWarnings = FALSE, recursive = TRUE)
```

## 1. Import files and preprocess data

### 1.1. Import count files (Kallisto)
We set the path and pattern for the Kallisto files to load the counts.

```{r import_counts}
files <- list.files(full.names = TRUE, 
                    path = "../data/counts", 
                    pattern = "*.tsv")
files
```

### 1.2. Import experimental metadata
We load the metadata, which contains the conditions and replicates for our experiment.

```{r import_metadata}
metadatos <- read.csv(file = "../data/metadata_all.csv", row.names = 1)
head(metadatos)
```

**Quality check:** We verify if the number of count files matches the number of samples in our metadata.
```{r check_match}
length(files) == nrow(metadatos)
```

### 1.3. Build the tx2gene object
Since we are working with bacteria (*Pectobacterium*), there is no complex alternative splicing. Because of this, the transcript name is the same as the gene name. 

*Note: We read the first count file to get the gene IDs without using absolute local paths.*

```{r build_tx2gene}
kallisto_ejemplo <- read.delim(files[1])

tx2gene <- data.frame(
  TXNAME = kallisto_ejemplo$target_id,
  GENEID = kallisto_ejemplo$target_id
)
```

### 1.4. Import with tximport and create DESeq2 object
We merge the counts and set up the `DESeqDataSet` object, using `Condition` as our main experimental variable.

```{r run_tximport}
txi.kallisto <- tximport(files, 
                         type = "kallisto", 
                         tx2gene = tx2gene) 

dds <- DESeqDataSetFromTximport(txi.kallisto, 
                                colData = metadatos, 
                                design = ~ Condition)
dds
```

### 1.5. Extract and filter counts
To improve our statistical results and reduce noise, we remove genes with low expression. We only keep genes that have an average of more than 10 counts across all samples.

```{r raw_and_filtered_counts}
# 1. Extract and save raw counts
rawcounts <- as.matrix(counts(dds))
head(rawcounts)
write.csv(rawcounts, file = "../results/rawcounts.csv", row.names = TRUE)

# 2. Filter low expression genes
filtcounts <- rawcounts[rowMeans(rawcounts) > 10, ]
head(filtcounts)

# 3. Calculate the percentage of genes kept
p <- (nrow(filtcounts) / nrow(rawcounts)) * 100
paste("Expressed genes kept after filtering =", round(p, 2), "%")

write.csv(filtcounts, file = "../results/filtcounts.csv", row.names = TRUE)
```

We update our `DESeqDataSet` object to only use the genes we kept.

```{r update_dds}
expgen <- rownames(filtcounts)
dds <- dds[expgen, ]
dds
```

---

## 2. Exploratory data analysis

### 2.1. Variance Stabilizing Transformation (VST)
We apply the VST to normalize the data. This makes the variance independent of the mean, which helps clustering methods like PCA work better.

```{r vst_transformation}
vsd <- vst(dds, blind = FALSE)
```

### 2.2. Principal Component Analysis (PCA)
We calculate and plot the PCA to see how similar the samples are and check if they group by their experimental condition.

```{r pca_calculation}
# Get coordinates and explained variance
pcaData <- plotPCA(vsd, intgroup = "Condition", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

paste("Total explained variance (PC1 + PC2) =", sum(percentVar), "%")
```

Create and save the PCA plot:

```{r pca_plot, fig.width=11, fig.height=8.5}
# Consistent palette: blue = Control, red = Treatment
pca_palette <- c("Control" = "#4575B4", "Treatment" = "#D73027")

pca <- ggplot(pcaData,
              aes(x = PC1, y = PC2,
                  color = Condition, shape = Condition)) +
  geom_point(size = 7, alpha = 0.85) +
  # circle (16) for Control, filled triangle (17) for Treatment — same in all PCAs
  scale_shape_manual(values = c("Control" = 16, "Treatment" = 17),
                     name = "Condition") +
  scale_color_manual(values = pca_palette, name = "Condition") +
  geom_text_repel(aes(label = name), show.legend = FALSE,
                  size = 4, fontface = "bold",
                  box.padding = 0.5, bg.color = "white", bg.r = 0.1) +
  labs(title    = "PCA of Pectobacterium, Treatment vs Control",
       subtitle  = "Filtered and VST normalized data",
       x = paste0("PC1: ", percentVar[1], "% variance"),
       y = paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 14)) +
  coord_fixed()

pca

# Save the plot in high quality
ggsave(filename = "../results/pca_raw.pdf", plot = pca, width = 11, height = 8.5)
ggsave(filename = file.path(tiff_dir, "pca_raw.tiff"), plot = pca, width = 11, height = 8.5, dpi = 300, compression = "lzw")
```

### 2.3. Save workspace
We save the objects in an `.RData` file so we can easily use them in future scripts without running everything again.

```{r save_rdata}
save(metadatos, dds, rawcounts, filtcounts, pca, file = "../results/dds.RData")
```

---

## Session information
Recording the R version and packages used helps make sure this analysis can be reproduced.

```{r session_info}
sessionInfo()
```