---
title: "Normalization and Batch Correction"
author: "López-Chablé Edgard Uriel"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: zenburn
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message = FALSE, warning = FALSE}
# Load necessary packages
library(tidyverse)
library(cowplot)
library(DESeq2)
library(sva)
library(ggrepel)

# TIFF output directory (created automatically)
tiff_dir <- "../results/TIFFs/2_DataNormalization"
dir.create(tiff_dir, showWarnings = FALSE, recursive = TRUE)
```

## 1. Import files

We load the workspace created in the previous exploratory analysis.
```{r load_rdata}
load(file = "../results/dds.RData")

```

We extract the batch information from the metadata. This is important to find and fix technical effects that are not related to the biology of the experiment.
```{r extract_batch}
head(metadatos)
batch <- metadatos$batch
batch
```

We check that the filtered counts object is still a matrix, which is needed for the correction functions.
```{r check_matrix_class}
class(filtcounts)
```

---

## 2. Batch Effect Correction

We use `ComBat_seq` to adjust the filtered raw counts. Unlike other tools, `ComBat_seq` keeps the counts as whole numbers, which is required by DESeq2.

```{r combat_seq_correction}
adjcounts <- ComBat_seq(filtcounts,      # Filtered counts matrix
                        batch = batch,   # Batch variable
                        group = NULL)    # No biological variables for now
head(adjcounts)
```

We save the adjusted counts matrix.
```{r export_adjcounts}
write.csv(adjcounts, file = "../results/adjcounts.csv", row.names = TRUE)
```

We update the `DESeqDataSet` object using the new batch-corrected matrix.
```{r create_dds_adj}
dds_adj <- DESeqDataSetFromMatrix(countData = adjcounts,
                                  colData = metadatos, 
                                  design = ~ Condition)
dds_adj
```

---

## 3. Exploratory analysis of adjusted counts

We apply a new Variance Stabilizing Transformation (VST) on the corrected data to see how well `ComBat_seq` worked using a PCA plot.

```{r pca_adj_calculation}
vsd2 <- vst(dds_adj, blind = FALSE)
pcaData2 <- plotPCA(vsd2, intgroup = "Condition", returnData = TRUE)

# Calculate explained variance
percentVar2 <- round(100 * attr(pcaData2, "percentVar"))
paste("Explained variance (PC1 + PC2) = ", sum(percentVar2), "%")
```

We create the PCA plot for the adjusted data.
```{r pca_adj_plot, fig.width=11, fig.height=8.5}
# Same palette and shapes as PCA in script 1
pca_palette <- c("Control" = "#4575B4", "Treatment" = "#D73027")

pca2 <- ggplot(pcaData2, aes(x = PC1, y = PC2,
                              color = Condition, shape = Condition)) +
  geom_point(size = 7, alpha = 0.85) +
  scale_shape_manual(values = c("Control" = 16, "Treatment" = 17),
                     name = "Condition") +
  scale_color_manual(values = pca_palette, name = "Condition") +
  geom_text_repel(aes(label = name), show.legend = FALSE,
                  size = 4, fontface = "bold",
                  box.padding = 0.5, bg.color = "white", bg.r = 0.1) +
  labs(title    = "PCA of Pectobacterium, Treatment vs Control",
       subtitle  = "Batch corrected data",
       x = paste0("PC1: ", percentVar2[1], "% variance"),
       y = paste0("PC2: ", percentVar2[2], "% variance")) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 14)) +
  coord_fixed()

pca2
ggsave(filename = "../results/pca_adj.pdf", plot = pca2, width = 11, height = 8.5)
ggsave(filename = file.path(tiff_dir, "pca_adj.tiff"), plot = pca2, width = 11, height = 8.5, dpi = 300, compression = "lzw")
```

We visually compare the original (raw) PCA with the corrected (adjusted) PCA.
```{r compare_pcas, fig.width=20, fig.height=4}
pcas <- plot_grid(pca, pca2,
                  ncol = 2, nrow = 1,
                  labels = "AUTO")
pcas

ggsave(filename = "../results/pca_raw_adj.pdf", plot = pcas, width = 20, height = 4)
ggsave(filename = file.path(tiff_dir, "pca_raw_adj.tiff"), plot = pcas, width = 20, height = 4, dpi = 300, compression = "lzw")
```

We save the progress of the adjusted data.
```{r save_adj_rdata}
save(dds_adj, adjcounts, pca2, pcas, file = "../results/dds_adj.RData")
```

---

## 4. Normalization for Differential Expression

DESeq2 uses the "median of ratios" method to calculate size factors. This fixes differences in sequencing depth between samples.

```{r deseq_normalization}
dds_norm <- DESeq(dds_adj)
dds_norm
```

Review of the size factors for each sample:
* Values `> 1` mean the sample was over-represented.
* Values `< 1` mean the sample was under-represented.
```{r check_size_factors}
dds_norm$sizeFactor
```

We extract and save the final normalized counts matrix.
```{r extract_normcounts}
normcounts <- counts(dds_norm, normalized = TRUE)
head(normcounts)
write.csv(normcounts, file = "../results/normcounts.csv", row.names = TRUE)
```

### 4.1. Evaluate data spread

To see the effect of our processing, we compare the counts in three stages: raw, batch-corrected, and normalized. We use a log transformation (log(x+1)) just to make the boxplots easier to read.

```{r log_transform}
log_raw <- log(filtcounts + 1)
log_adj <- log(adjcounts + 1)
log_norm <- log(normcounts + 1)
```

Helper function to change the matrices from wide format to long format. This makes them work with `ggplot2`.
```{r wide2long_function}
wide2long <- function(countsmatrix) {
    as.data.frame(countsmatrix) %>% 
    rownames_to_column("Gene") %>%  
    pivot_longer(cols = -Gene,      
                 names_to = "Sample", 
                 values_to = "Count") 
}
```

We apply this function to our three matrices:
```{r reshape_matrices}
log_raw_long <- wide2long(log_raw)
log_adj_long <- wide2long(log_adj) 
log_norm_long <- wide2long(log_norm) 
```

Create comparative boxplots:
```{r generate_boxplots}
# 1. Raw counts boxplot
box1 <- ggplot(log_raw_long, aes(x = Sample, y = Count, fill = Sample)) + 
    geom_boxplot() + 
    labs(title = "Raw filtered counts", 
         x = "Samples", 
         y = "log(counts)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1), 
          legend.position = "none") 

# 2. Batch corrected boxplot
box2 <- ggplot(log_adj_long, aes(x = Sample, y = Count, fill = Sample)) +
    geom_boxplot() +
    labs(title = "Batch corrected counts", 
         x = "Samples", 
         y = "log(counts)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1), 
          legend.position = "none")

# 3. Normalized counts boxplot
box3 <- ggplot(log_norm_long, aes(x = Sample, y = Count, fill = Sample)) +
    geom_boxplot() +
    labs(title = "Normalized counts (DESeq2)", 
         x = "Samples", 
         y = "log(counts)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1), 
          legend.position = "none")
```

We group and save the final plots.
```{r combine_boxplots, fig.width=11, fig.height=8.5}
boxplots <- plot_grid(box1, box2, box3,
                      ncol = 3, nrow = 1,
                      labels = "AUTO")
boxplots

ggsave(filename = "../results/boxplots.pdf", plot = boxplots, width = 15, height = 6)
ggsave(filename = file.path(tiff_dir, "boxplots.tiff"), plot = boxplots, width = 15, height = 6, dpi = 300, compression = "lzw")
```

We save the final normalized objects for the differential expression analysis.
```{r save_final_rdata}
save(dds_norm, normcounts, file = "../results/dds_norm.RData")
```

---

## Session information
```{r session_info}
sessionInfo()
```