---
title: "Functional Enrichment Analysis of DEGs"
author: "López-Chablé Edgard Uriel"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: zenburn
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message = FALSE, warning = FALSE}
# Load all required packages
library(clusterProfiler)
library(enrichplot)
library(dplyr)
library(tidyr) 
library(ggplot2)
library(pathview)
library(GO.db)
library(readr)
library(stringr)
library(UpSetR)
library(STRINGdb)
library(magick)

# TIFF output directory (created automatically)
tiff_dir <- "../results/TIFFs/4_FunctionalCategorization"
dir.create(tiff_dir, showWarnings = FALSE, recursive = TRUE)
```

## 1. Setup environment and directories

We create folders to organize our functional enrichment outputs.
```{r create_directories}
out_dir <- "../results/enrichment"
kegg_dir <- "../results/enrichment/kegg"
adv_dir <- "../results/advanced_analysis"

dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(kegg_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(adv_dir, showWarnings = FALSE, recursive = TRUE)

```

## 2. Import and prepare data

### 2.1. Load functional metadata (eggNOG)
We load the eggNOG annotation and build the `TERM2GENE` objects to map genes to their functions.

```{r load_eggnog}
# Load base annotation
eggnog_data <- read.delim("../data/annotation.tsv")

# GO Dictionary
term2gene_go <- eggnog_data %>%
  dplyr::select(GOs, X.query) %>%
  tidyr::separate_rows(GOs, sep = ",") %>%
  dplyr::filter(GOs != "" & GOs != "-" & !is.na(GOs)) %>%
  dplyr::mutate(query = trimws(X.query), GOs = trimws(GOs)) %>% 
  dplyr::select(GOs, X.query) %>%
  as.data.frame()

# Map GO IDs to descriptive terms
vector_go <- Term(GOTERM)
term2name_go <- data.frame(
  GO_ID = as.character(names(vector_go)),
  Term = as.character(unname(vector_go)),
  stringsAsFactors = FALSE
)

# KEGG Dictionary (Filtered for 'map' metabolic pathways)
term2gene_kegg <- eggnog_data %>%
  dplyr::select(KEGG_Pathway, X.query) %>%
  tidyr::separate_rows(KEGG_Pathway, sep = ",") %>%
  dplyr::filter(KEGG_Pathway != "" & KEGG_Pathway != "-" & !is.na(KEGG_Pathway)) %>%
  dplyr::filter(grepl("map", KEGG_Pathway)) %>% 
  dplyr::mutate(query = trimws(X.query), KEGG_Pathway = trimws(KEGG_Pathway)) %>%
  dplyr::select(KEGG_Pathway, X.query) %>%
  as.data.frame()
```

### 2.2. Import Differentially Expressed Genes (DEGs)
We extract and separate the significant genes found in our previous analysis.

```{r load_degs, warning=FALSE}
# Import significant results and clean IDs
deg_table <- read_csv("../results/deg_Treatment_vs_Control.csv") %>% 
  mutate(GeneID = str_extract(...1, "WP_[0-9]+\\.[0-9]+"))

# Separate genes by regulation direction
up_genes <- deg_table %>% filter(log2FoldChange > 0) %>% pull(GeneID) %>% trimws()
down_genes <- deg_table %>% filter(log2FoldChange < 0) %>% pull(GeneID) %>% trimws()
all_deg_ids <- trimws(deg_table$GeneID)
```

---

## 3. Enrichment Analysis (ORA)

### 3.1. Gene Ontology (GO)
We test for over-represented biological processes in both up-regulated and down-regulated genes.

```{r go_enrichment}
# ORA: Up-regulated
enrich_go_up <- enricher(gene = up_genes, 
                         TERM2GENE = term2gene_go, 
                         TERM2NAME = term2name_go,
                         pAdjustMethod = "BH", 
                         pvalueCutoff  = 0.1, 
                         qvalueCutoff  = 0.1)

# ORA: Down-regulated
enrich_go_down <- enricher(gene = down_genes, 
                           TERM2GENE = term2gene_go, 
                           TERM2NAME = term2name_go,
                           pAdjustMethod = "BH", 
                           pvalueCutoff  = 0.1, 
                           qvalueCutoff  = 0.1)
```

Create and export the Dotplots.
```{r go_dotplots, fig.width=10, fig.height=7}
p_go_up <- enrichplot::dotplot(enrich_go_up, showCategory = 15,
                               title = "GO Enrichment (Up-regulated)") +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 45)) +
  scale_color_gradient(low = "#4575B4", high = "#D73027",
                       name = "adj. p-value") +
  theme_minimal(base_size = 12) +
  theme(plot.title   = element_text(face = "bold", size = 14),
        axis.text.y  = element_text(size = 10),
        axis.text.x  = element_text(size = 10),
        legend.title = element_text(size = 10))

p_go_down <- enrichplot::dotplot(enrich_go_down, showCategory = 15,
                                 title = "GO Enrichment (Down-regulated)") +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 45)) +
  scale_color_gradient(low = "#4575B4", high = "#D73027",
                       name = "adj. p-value") +
  theme_minimal(base_size = 12) +
  theme(plot.title   = element_text(face = "bold", size = 14),
        axis.text.y  = element_text(size = 10),
        axis.text.x  = element_text(size = 10),
        legend.title = element_text(size = 10))

p_go_up
p_go_down

ggsave(filename = paste0(out_dir, "/dotplot_GO_UP.pdf"),   plot = p_go_up,   width = 12, height = 8)
ggsave(filename = file.path(tiff_dir, "dotplot_GO_UP.tiff"),   plot = p_go_up,   width = 12, height = 8, dpi = 300, compression = "lzw")
ggsave(filename = paste0(out_dir, "/dotplot_GO_DOWN.pdf"), plot = p_go_down, width = 12, height = 8)
ggsave(filename = file.path(tiff_dir, "dotplot_GO_DOWN.tiff"), plot = p_go_down, width = 12, height = 8, dpi = 300, compression = "lzw")
```

### 3.2. KEGG Pathway Enrichment
We check for affected metabolic pathways.

```{r kegg_enrichment, fig.width=10, fig.height=7}
enrich_kegg <- enricher(gene = all_deg_ids, 
                        TERM2GENE = term2gene_kegg, 
                        pAdjustMethod = "BH", 
                        pvalueCutoff  = 0.1, 
                        qvalueCutoff  = 0.1)

p_kegg <- enrichplot::dotplot(enrich_kegg, showCategory = 15,
                              title = "KEGG Pathway Enrichment") +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 45)) +
  scale_color_gradient(low = "#4575B4", high = "#D73027",
                       name = "adj. p-value") +
  theme_minimal(base_size = 12) +
  theme(plot.title   = element_text(face = "bold", size = 14),
        axis.text.y  = element_text(size = 10),
        axis.text.x  = element_text(size = 10),
        legend.title = element_text(size = 10))

p_kegg
ggsave(filename = paste0(out_dir, "/dotplot_KEGG.pdf"),  plot = p_kegg, width = 12, height = 8)
ggsave(filename = file.path(tiff_dir, "dotplot_KEGG.tiff"),  plot = p_kegg, width = 12, height = 8, dpi = 300, compression = "lzw")
```

---

## 4. Automated KEGG Pathway Visualization (Pathview)

We download the top 15 affected metabolic pathways and color them using our `log2FoldChange` values.

```{r prepare_pathview}
# Get KEGG Orthology (KO) IDs
eggnog_ko <- eggnog_data %>%
  dplyr::select(X.query, KEGG_ko) %>%
  dplyr::filter(KEGG_ko != "" & KEGG_ko != "-") %>%
  dplyr::mutate(KEGG_ko = gsub("ko:", "", KEGG_ko)) %>%
  dplyr::mutate(KEGG_ko = sapply(strsplit(as.character(KEGG_ko), ","), `[`, 1))

# Join with Fold Changes
pathview_data <- deg_table %>%
  dplyr::inner_join(eggnog_ko, by = c("GeneID" = "X.query")) %>%
  dplyr::filter(!is.na(KEGG_ko) & !is.na(log2FoldChange))

# Create a numeric vector for expression values
fc_vector <- pathview_data$log2FoldChange
names(fc_vector) <- pathview_data$KEGG_ko
```

Run the rendering loop.
```{r render_pathview_loop, results='hide', message=FALSE, warning=FALSE}
# Extract top 15 KEGG IDs and clean the prefix
top_kegg_ids <- head(enrich_kegg@result$ID, 15)
top_kegg_clean <- gsub("map", "", top_kegg_ids)

# Temporarily change directory to save outputs neatly
current_wd <- getwd()
setwd(kegg_dir) 

# 1. Loop to download and render each map (saved as PNGs)
for (path_id in top_kegg_clean) {
  message("Processing KEGG pathway: ", path_id)
  
  tryCatch({
    # Round limit to nearest integer so legend labels don't overlap
    fc_limit <- ceiling(max(abs(fc_vector)))
    pathview(gene.data  = fc_vector,
             pathway.id = path_id,
             species    = "ko",
             limit      = list(gene = fc_limit, cpd = 1),
             kegg.native = TRUE) 
  }, error = function(e) {
    message("Skipped pathway ", path_id, " (Insufficient data or API constraint).")
  })
}

# 2. Combine all maps into a single PDF safely
library(magick)

# Get all generated PNGs
png_files <- list.files(pattern = "\\.pathview\\.png$")

# SAFETY FILTER: Keep only files larger than 5KB to avoid "ImproperImageHeader" errors
# from failed KEGG downloads (which are usually 0 bytes or tiny HTML error files).
valid_png_files <- png_files[file.info(png_files)$size > 5000] 

if(length(valid_png_files) > 0) {
  message("Combining ", length(valid_png_files), " valid maps into a single PDF...")
  
  # Read only the valid, uncorrupted images
  kegg_maps <- image_read(valid_png_files)
  
  # Write to a single PDF
  image_write(kegg_maps, format = "pdf", path = "Top15_KEGG_Pathways_Consolidated.pdf")
  
  # Clean up all downloaded PNGs (both valid and corrupted) to keep the folder clean
  file.remove(png_files) 
} else {
  message("No valid maps were generated to combine.")
  # Clean up any leftover corrupted files
  file.remove(png_files)
}

# Return to original working directory
setwd(current_wd)
```

---

## 5. Network Analysis (PPI - STRINGdb)
We search for functional networks using orthologs confirmed by eggNOG.

```{r string_network, warning=FALSE}
# Significant genes (padj < 0.05) with a seed ortholog, capped at the 1001 lowest adjusted p-values
ppi_data <- deg_table %>%
  dplyr::inner_join(eggnog_data, by = c("GeneID" = "X.query")) %>%
  dplyr::filter(padj < 0.05 & seed_ortholog != "") %>%
  dplyr::arrange(padj) %>%
  dplyr::slice_head(n = 1001) 

string_ids <- ppi_data$seed_ortholog
tax_id <- as.numeric(strsplit(string_ids[1], "\\.")[[1]][1])

string_db <- STRINGdb$new(version = "11.5", 
                          species = tax_id, 
                          score_threshold = 700, 
                          input_directory = out_dir)

string_db$plot_network(string_ids)
```

---

## 6. Specialized Plots

### 6.1. UpSet Plot (Intersections)
```{r upset_plot, fig.width=8, fig.height=6}
intersect_list <- list(
  "Up-regulated" = up_genes,
  "Down-regulated" = down_genes
)

p_upset <- upset(fromList(intersect_list), 
                 order.by = "freq", 
                 mainbar.y.label = "Intersection Size", 
                 sets.x.label = "Total Genes",
                 text.scale = c(1.3, 1.3, 1, 1, 1.5, 1.2), 
                 point.size = 3.5, 
                 line.size = 1)

print(p_upset)
```

### 6.2. Enrichment Map
We use `emapplot` with a Fruchterman-Reingold layout (`fr`) which spreads similar terms
more than `nicely`. Keeping only the top 12 terms avoids the central hairball typical when
all enriched GO terms belong to the same super-family (e.g. purine/ATP biosynthesis).

```{r emap_plots, fig.width=10, fig.height=9}
# ── Up-regulated ──────────────────────────────────────────────────────────────
enrich_go_up_sim <- pairwise_termsim(enrich_go_up)

p_emap_up <- emapplot(enrich_go_up_sim,
                      showCategory = 12,
                      color        = "p.adjust",
                      layout       = "fr") +   # Fruchterman-Reingold spreads clusters better
  scale_color_gradient(low = "#1A9641", high = "#D73027",
                       name = "adj. p-value") +
  ggtitle("Enrichment Map — Up-regulated GO Terms") +
  theme(plot.title  = element_text(face = "bold", size = 14, hjust = 0.5),
        legend.text = element_text(size = 9))

p_emap_up
ggsave(filename = paste0(out_dir, "/emapplot_UP.pdf"),
       plot = p_emap_up, width = 11, height = 9)
ggsave(filename = file.path(tiff_dir, "emapplot_UP.tiff"),
       plot = p_emap_up, width = 11, height = 9, dpi = 300, compression = "lzw")

# ── Down-regulated ────────────────────────────────────────────────────────────
if (!is.null(enrich_go_down) && nrow(enrich_go_down@result) > 1) {
  enrich_go_down_sim <- pairwise_termsim(enrich_go_down)

  p_emap_down <- emapplot(enrich_go_down_sim,
                          showCategory = 12,
                          color        = "p.adjust",
                          layout       = "fr") +
    scale_color_gradient(low = "#1A9641", high = "#D73027",
                         name = "adj. p-value") +
    ggtitle("Enrichment Map — Down-regulated GO Terms") +
    theme(plot.title  = element_text(face = "bold", size = 14, hjust = 0.5),
          legend.text = element_text(size = 9))

  p_emap_down
  ggsave(filename = paste0(out_dir, "/emapplot_DOWN.pdf"),
         plot = p_emap_down, width = 11, height = 9)
  ggsave(filename = file.path(tiff_dir, "emapplot_DOWN.tiff"),
         plot = p_emap_down, width = 11, height = 9, dpi = 300, compression = "lzw")
} else {
  message("Not enough down-regulated GO terms to build an enrichment map.")
}
```

---

## 7. Prepare Data for Advanced Analysis (GSEA)

To run tests like Gene Set Enrichment Analysis (GSEA) or upstream regulator checks, we need the full dataset, not just the filtered significant genes.

```{r prep_advanced_data, warning=FALSE, message=FALSE}
# Import the full DESeq2 results (no p-value filter)
full_res <- read_csv("../results/deseq_results_Treatment_vs_Control.csv") %>%
  mutate(GeneID = str_extract(...1, "WP_[0-9]+\\.[0-9]+")) %>%
  filter(!is.na(GeneID) & !is.na(log2FoldChange))

# Create a ranked list required for GSEA
ranked_list <- full_res %>%
  arrange(desc(log2FoldChange)) %>%
  dplyr::select(GeneID, log2FoldChange)

# Join the full data with eggNOG annotations
full_annotation <- full_res %>%
  left_join(eggnog_data, by = c("GeneID" = "X.query"))

# Save objects for the next pipeline
saveRDS(ranked_list, file = paste0(adv_dir, "/gsea_ranked_list.rds"))
write_csv(full_annotation, file = paste0(adv_dir, "/full_annotated_results.csv"))
```

---

## Session information
```{r session_info}
sessionInfo()
```