---
title: "Advanced Analysis: GSEA, Virulence, and Networks"
author: "López-Chablé Edgard Uriel"
output: 
  pdf_document:
    toc: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: zenburn
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message = FALSE, warning = FALSE}
# Load necessary packages
library(clusterProfiler) # For enrichment analysis (GO/KEGG)
library(enrichplot)      # To plot enrichment results
library(tidyverse)       # For organizing and cleaning data
library(ggplot2)         # For custom plots
library(pheatmap)        # To create heatmaps
library(viridis)         # Color palettes
library(ggrepel)         # To prevent text overlap in plots
library(KEGGREST)        # To get pathway names from the KEGG API
library(STRINGdb)        # To get protein-protein interactions
library(igraph)          # To build and analyze networks
library(ggraph)          # To graph networks
library(tidygraph)       # To modify networks

# TIFF output directory (created automatically)
tiff_dir <- "../results/TIFFs/5_AdvancedAnalysis"
dir.create(tiff_dir, showWarnings = FALSE, recursive = TRUE)
```

## 1. Environment Setup

We create folders to keep things organized. Analyses like KEGG create many images that can quickly clutter the main folder.

```{r create_directories}
adv_dir <- "../results/advanced_analysis"
dir.create(adv_dir, showWarnings = FALSE, recursive = TRUE)

```

## 2. Data Import

We load the ranked list (ordered by Log2FC) and the full database. GSEA needs the complete list of genes, not just the significant ones.

```{r load_data}
# Prepare the ranked list for GSEA 
ranked_df <- readRDS("../results/advanced_analysis/gsea_ranked_list.rds") %>%
  # Keep only the highest log2FC value for duplicate genes
  group_by(GeneID) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

gsea_vector <- ranked_df$log2FoldChange
names(gsea_vector) <- ranked_df$GeneID
gsea_vector <- sort(gsea_vector, decreasing = TRUE)

# Load full differential expression data
full_annotation <- read_csv("../results/advanced_analysis/full_annotated_results.csv")

# Rebuild the KEGG dictionary from our eggNOG file
eggnog_data <- read.delim("../data/annotation.tsv")

term2gene_kegg <- eggnog_data %>%
  dplyr::select(KEGG_Pathway, X.query) %>%
  tidyr::separate_rows(KEGG_Pathway, sep = ",") %>%
  dplyr::filter(grepl("map", KEGG_Pathway)) %>% 
  dplyr::mutate(query = trimws(X.query), KEGG_Pathway = trimws(KEGG_Pathway)) %>%
  dplyr::select(KEGG_Pathway, X.query) %>%
  as.data.frame()
```

---

## 3. Gene Set Enrichment Analysis (GSEA)

GSEA looks at the whole ordered gene list to find if entire pathways are globally up-regulated or down-regulated.

```{r run_gsea, warning=FALSE, message=FALSE}
# Get official KEGG pathway names from the web
kegg_names <- keggList("pathway")

# Clean up the names to make them easier to read
term2name_kegg <- data.frame(
  PathwayID = sub("path:", "", names(kegg_names)), 
  PathwayName = as.character(kegg_names)
)
term2name_kegg$PathwayName <- gsub(" - Reference pathway", "", term2name_kegg$PathwayName)

# Run GSEA (seeded so that p-values are reproducible)
set.seed(124)
gsea_kegg <- GSEA(geneList = gsea_vector, 
                  TERM2GENE = term2gene_kegg, 
                  TERM2NAME = term2name_kegg, 
                  pvalueCutoff = 0.05, 
                  pAdjustMethod = "BH",
                  minGSSize = 10,  
                  maxGSSize = 500, 
                  eps = 0,
                  seed = TRUE)   # reproducible permutations (uses set.seed above)
```

### 3.1. Systemic Shifts (Ridgeplot)
This plot shows how genes in the significant pathways are distributed.

```{r ridgeplot, fig.width=10, fig.height=8}
p_ridge <- ridgeplot(gsea_kegg, showCategory = 10) +
  labs(title = "GSEA Ridgeplot: Systemic KEGG Pathway Shifts",
       x = "Enrichment Distribution (Log2FC)") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16))

p_ridge
ggsave(filename = paste0(adv_dir, "/gsea_ridgeplot.pdf"),  plot = p_ridge, width = 10, height = 8)
ggsave(filename = file.path(tiff_dir, "gsea_ridgeplot.tiff"), plot = p_ridge, width = 10, height = 8, dpi = 300, compression = "lzw")
```

### 3.2. Top Enriched Pathway Plot

```{r gsea_plot, fig.width=10, fig.height=8.5}
# Pick a specific pathway (e.g., the second one) to avoid generic ones like Ribosome
target_pathway_idx <- 2 
top_gsea_pathway <- gsea_kegg@result$ID[target_pathway_idx] 

p_gseaplot <- gseaplot2(gsea_kegg, 
                        geneSetID = top_gsea_pathway, 
                        title = paste("GSEA - Pathway:", gsea_kegg@result$Description[target_pathway_idx]), 
                        pvalue_table = TRUE)

pdf(paste0(adv_dir, "/target_pathway_gseaplot.pdf"), width = 10, height = 8.5)
print(p_gseaplot)
dev.off()
tiff(file.path(tiff_dir, "target_pathway_gseaplot.tiff"), width = 10, height = 8.5,
     units = "in", res = 300, compression = "lzw")
print(p_gseaplot)
dev.off()
```

---

## 4. Virulence Profiling

We filter the data to find plant cell wall degrading enzymes and secretion systems to see how the pathogen's attack tools change.

```{r extract_virulence}
# Keywords to find virulence factors.
# Functional keywords are searched in the eggNOG description; the gene-family
# prefixes (pel, peh, out, kdg) are matched only at the start of the gene name,
# so that e.g. "out" does not capture every "outer membrane" protein.
virulence_desc_kw <- "pectate lyase|cellulase|protease|secretion system"
virulence_name_kw <- "^(pel|peh|out|kdg)"

virulence_df <- full_annotation %>%
  filter(padj < 0.05) %>% 
  filter(str_detect(tolower(Description), virulence_desc_kw) | 
         str_detect(tolower(Preferred_name), virulence_name_kw)) %>%
  arrange(desc(abs(log2FoldChange)))

virulence_genes <- virulence_df$GeneID
message("Found ", length(virulence_genes), " significant virulence factors.")

write_csv(virulence_df, paste0(adv_dir, "/virulence_factors_table.csv"))
```

### 4.1. Virulence Factors Heatmap

```{r virulence_heatmap, fig.width=8, fig.height=10}
# Load normalized counts and metadata
load(file = "../results/dds_norm.RData") 
metadatos <- read.csv(file = "../data/metadata_all.csv", row.names = 1)

colnames(normcounts) <- metadatos$Replicates

# Fix row names to match our GeneIDs
rownames(normcounts) <- str_extract(rownames(normcounts), "WP_[0-9]+\\.[0-9]+")
normcounts <- normcounts[!is.na(rownames(normcounts)), ]

# Keep only the virulence genes
virulence_matrix <- normcounts[virulence_genes, ]

# Use common gene names (like pelA) if we have them
rownames(virulence_matrix) <- ifelse(!is.na(virulence_df$Preferred_name), 
                                     virulence_df$Preferred_name, 
                                     virulence_df$GeneID)

p_virulence_hm <- pheatmap(virulence_matrix, 
                           scale = "row", 
                           color = viridis(100),
                           treeheight_row = 15,
                           treeheight_col = 15,
                           main = "Pectobacterium Virulence Factors Expression",
                           fontsize_row = 8)

ggsave(filename = paste0(adv_dir, "/heatmap_virulence.pdf"),  plot = p_virulence_hm, width = 8, height = 10)
ggsave(filename = file.path(tiff_dir, "heatmap_virulence.tiff"), plot = p_virulence_hm, width = 8, height = 10, dpi = 300, compression = "lzw")
```

---

## 5. Upstream Regulators Analysis

We search for transcription factors and DNA-binding proteins that might be controlling these changes.

```{r tf_volcano, fig.width=10, fig.height=8, warning=FALSE}
tf_keywords <- "transcription factor|regulator|dna-binding"   # matched against lower-cased text

# Four-category coloring: background / NS regulator / up-regulator / down-regulator
volcano_data <- full_annotation %>%
  mutate(
    Is_Regulator = if_else(
      str_detect(tolower(Description), tf_keywords) |
        str_detect(tolower(Preferred_name), tf_keywords), "Yes", "No"),
    Category = case_when(
      Is_Regulator == "Yes" & !is.na(padj) & padj < 0.05 & log2FoldChange >  1 ~ "Regulator ↑ in Treatment",
      Is_Regulator == "Yes" & !is.na(padj) & padj < 0.05 & log2FoldChange < -1 ~ "Regulator ↓ in Treatment",
      Is_Regulator == "Yes"                                                      ~ "Regulator (NS)",
      TRUE                                                                       ~ "Background"
    ),
    Label = if_else(
      Is_Regulator == "Yes" & !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1,
      coalesce(
        na_if(Preferred_name, "-"),          # use Preferred_name when available
        str_extract(GeneID, "WP_[0-9]+\\.[0-9]+"),  # else short WP_ id
        GeneID                               # last resort: full id
      ),
      NA_character_)                         # NA so ggrepel truly ignores non-labeled points
  )

# Palette: red = higher in Treatment (pos FC), blue = higher in Control (neg FC)
reg_colors <- c("Background"              = "#BDBDBD",
                "Regulator (NS)"          = "#9ECAE1",
                "Regulator ↑ in Treatment"   = "#D73027",
                "Regulator ↓ in Treatment" = "#4575B4")
reg_sizes  <- c("Background"=1.2, "Regulator (NS)"=2.5,
                "Regulator ↑ in Treatment"=3.5, "Regulator ↓ in Treatment"=3.5)
reg_alpha  <- c("Background"=0.35, "Regulator (NS)"=0.65,
                "Regulator ↑ in Treatment"=1, "Regulator ↓ in Treatment"=1)

p_tf_volcano <- ggplot(volcano_data, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = Category, size = Category, alpha = Category)) +
  scale_color_manual(values = reg_colors, name = NULL) +
  scale_size_manual( values = reg_sizes,  name = NULL) +
  scale_alpha_manual(values = reg_alpha,  name = NULL) +
  geom_text_repel(data = ~ filter(.x, !is.na(Label)),
                  aes(label = Label),
                  size = 3.5, fontface = "bold",
                  box.padding = 0.5, point.padding = 0.3,
                  max.overlaps = Inf,
                  color = "black", bg.color = "white", bg.r = 0.12,
                  segment.color = "grey50", segment.size = 0.4) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed",
             color = "grey30", alpha = 0.6, linewidth = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",
             color = "grey30", alpha = 0.6, linewidth = 0.5) +
  labs(title    = "Global Transcriptional Regulators",
       subtitle = "Red = higher in Treatment (log2FC > 0) · Blue = higher in Control (log2FC < 0)",
       x = "log₂ Fold Change  (Treatment / Control)",
       y = "-log₁₀(adj. p-value)") +
  theme_minimal(base_size = 13) +
  theme(plot.title       = element_text(face = "bold", size = 15),
        plot.subtitle    = element_text(size = 10, color = "grey40"),
        panel.grid.minor = element_blank(),
        legend.position  = "right")

p_tf_volcano
ggsave(filename = paste0(adv_dir, "/volcano_regulators.pdf"),  plot = p_tf_volcano, width = 10, height = 8)
ggsave(filename = file.path(tiff_dir, "volcano_regulators.tiff"), plot = p_tf_volcano, width = 10, height = 8, dpi = 300, compression = "lzw")
```

---

## 6. Bioenergetics and Energy Targets

We look for genes related to energy, like pumps and electron transport, which can be useful for lab experiments.

```{r electrochemistry_targets}
electro_keywords <- "atp synthase|efflux|pump|ion channel|cytochrome|electron transport|nadh dehydrogenase|proton|potassium|sodium|calcium|voltage"

electro_genes <- full_annotation %>%
  filter(padj < 0.05) %>% 
  filter(str_detect(tolower(Description), electro_keywords) | 
         str_detect(tolower(Preferred_name), electro_keywords) |
         str_detect(tolower(GOs), "ion transport|atp biosynthetic")) %>% 
  arrange(desc(log2FoldChange)) 

message("Found ", nrow(electro_genes), " bioenergetic targets.")

write_csv(electro_genes, paste0(adv_dir, "/electrochemistry_targets.csv"))
```

---

## 7. Network Analysis (PPI Hubs & Clusters)

We build a Protein-Protein Interaction (PPI) network using STRING's mapping tool. We apply a High-Confidence threshold (Score > 700) to remove noise and isolate distinct functional modules.

```{r ppi_topology, warning=FALSE, message=FALSE}
# 1. Prepare data for STRING mapping
deg_network <- full_annotation %>%
  filter(padj < 0.05) %>%
  dplyr::select(GeneID, Preferred_name, seed_ortholog, log2FoldChange) %>%
  mutate(search_name = coalesce(Preferred_name, GeneID)) %>%
  group_by(search_name) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

# 2. Get Taxonomy ID and start STRINGdb (HIGH CONFIDENCE: 700)
valid_orthologs <- deg_network$seed_ortholog[!is.na(deg_network$seed_ortholog) & deg_network$seed_ortholog != ""]
tax_id <- as.numeric(strsplit(valid_orthologs[1], "\\.")[[1]][1])

string_db <- STRINGdb$new(version = "11.5", species = tax_id, score_threshold = 700)

# 3. Map genes securely
mapped_genes <- string_db$map(as.data.frame(deg_network), "search_name", removeUnmappedRows = TRUE)

# 4. Get interactions
interactions <- string_db$get_interactions(mapped_genes$STRING_id)

# 5. Build and clean the network graph
ppi_graph <- graph_from_data_frame(d = interactions, directed = FALSE)
ppi_graph <- simplify(ppi_graph)

# 6. Process topology and extract Bottleneck Hubs
if (vcount(ppi_graph) > 0) {
  # Find clusters
  # Louvain is stochastic: without a seed, module membership (and therefore the
  # per-module hub selection below) varies between runs. Seed 124 reproduces the
  # hub set shown in the published figure; 13 of the 15 hubs are stable across seeds.
  set.seed(124)
  communities <- cluster_louvain(ppi_graph)
  ppi_graph <- set_vertex_attr(ppi_graph, "cluster", value = as.numeric(membership(communities)))

  # Calculate importance metrics
  ppi_graph <- set_vertex_attr(ppi_graph, "degree", value = degree(ppi_graph))
  ppi_graph <- set_vertex_attr(ppi_graph, "betweenness", value = betweenness(ppi_graph))

  # Add names back to nodes
  node_metadata <- data.frame(STRING_id = V(ppi_graph)$name) %>%
    left_join(mapped_genes, by = "STRING_id") %>%
    distinct(STRING_id, .keep_all = TRUE)

  ppi_graph <- set_vertex_attr(ppi_graph, "label", value = node_metadata$search_name)
  ppi_graph <- set_vertex_attr(ppi_graph, "log2FC", value = node_metadata$log2FoldChange)

  # STRATEGY CHANGE: Extract hubs based on Betweenness (Bottlenecks) 
  # This labels critical bridges connecting different biological pathways.
  top_hubs <- node_metadata %>%
    mutate(Degree = V(ppi_graph)$degree, 
           Betweenness = V(ppi_graph)$betweenness, 
           Cluster = V(ppi_graph)$cluster) %>%
    group_by(Cluster) %>%
    slice_max(order_by = Betweenness, n = 3, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(desc(Betweenness)) %>%
    head(15)

  write_csv(top_hubs, paste0(adv_dir, "/ppi_bottleneck_hubs.csv"))
} else {
  message("STRINGdb could not find any interactions above the 700 threshold.")
}
```

### 7.1. Network Visualization (ggraph)
We use `ggraph` with a Kamada-Kawai layout to separate distinct functional modules, while
`ggrepel` highlights the bottleneck hubs identified above.

```{r plot_network, fig.width=12, fig.height=12, warning=FALSE}
library(ggraph)
library(tidygraph)
library(magick)

if (vcount(ppi_graph) > 0) {
  net_components <- components(ppi_graph)
  if (max(net_components$csize) > 1) {
    main_component <- induced_subgraph(ppi_graph, vids = which(net_components$membership == which.max(net_components$csize)))
    tidy_net <- as_tbl_graph(main_component) %>%
      mutate(
        is_hub            = name %in% top_hubs$STRING_id,
        hub_label         = ifelse(is_hub, label, NA),
        clean_betweenness = replace_na(betweenness, 0),
        max_bet           = max(clean_betweenness, na.rm = TRUE),
        node_size         = ifelse(max_bet > 0, (clean_betweenness / max_bet) * 10 + 3, 5)
      )

    p_net <- ggraph(tidy_net, layout = "kk") +
      geom_edge_link(color = "grey85", width = 0.5, alpha = 0.6) +
      geom_node_point(aes(size = node_size, color = as.factor(cluster), alpha = is_hub)) +
      scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.4), guide = "none") +
      geom_node_text(aes(label = hub_label),
                     repel         = TRUE,
                     fontface      = "bold",
                     size          = 5,
                     box.padding   = 1,
                     point.padding = 0.5,
                     max.overlaps  = Inf,
                     bg.color      = "white",
                     bg.r          = 0.15,
                     segment.color = "grey40",
                     segment.size  = 0.6) +
      scale_color_viridis_d(option = "viridis") +
      scale_size_continuous(range = c(1.5, 11), guide = "none") +
      labs(title    = "Protein-Protein Interaction Network",
           subtitle = "Top 15 Bottleneck Hubs Highlighted (High Confidence STRING Score > 700)") +
      theme_void() +
      theme(legend.position = "none",
            plot.title    = element_text(face = "bold", size = 18, hjust = 0.5),
            plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40"))

    # Render ONCE to PDF, then convert to TIFF with magick (avoids triple-rendering)
    pdf_path  <- paste0(adv_dir, "/ppi_high_confidence_network.pdf")
    tiff_path <- file.path(tiff_dir, "ppi_high_confidence_network.tiff")

    pdf(pdf_path, width = 12, height = 12)
    print(p_net)
    dev.off()

    image_write(
      image_read_pdf(pdf_path, density = 300),
      path = tiff_path, format = "tiff", compression = "lzw"
    )

    print(p_net)

  } else {
    message("No connected components found to draw the plot.")
  }
} else {
  message("Plot cancelled: the network is empty.")
}
```


---

## Session Information
```{r session_info}
sessionInfo()
```
