---
title: "Paper Figures (High-Resolution, 1200 dpi)"
author: "López-Chablé Edgard Uriel"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: readable
    highlight: zenburn
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 0. Purpose

This script regenerates the final set of figures selected for the manuscript, exported at
**1200 dpi** into a dedicated `results/paper_images` folder. It reuses the objects already
computed by scripts `3_DifferentialExpression.Rmd`, `4_FunctionalCategorization.Rmd`, and
`5_AdvancedAnalysis.Rmd` — run those first if the intermediate files listed below are missing.

Figures produced:

1. Volcano plot — all genes (Treatment vs Control)
2. KEGG pathway enrichment (GSEA ridgeplot, "Photosynthesis" pathways excluded)
3. Protein-protein interaction (PPI) network, high confidence (STRING score > 700)
4. Volcano plot — virulence-associated genes highlighted
5. Heatmap — virulence-associated genes
6. Volcano plot — transcriptional regulators highlighted

```{r libraries, message = FALSE, warning = FALSE}
library(DESeq2)
library(EnhancedVolcano)
library(tidyverse)
library(pheatmap)
library(viridis)
library(ggrepel)
library(clusterProfiler)
library(enrichplot)
library(KEGGREST)
library(STRINGdb)
library(igraph)
library(ggraph)
library(tidygraph)
library(magick)
library(pdftools)

# Dedicated output folder for final paper figures (1200 dpi)
paper_dir <- "../results/paper_images"
dir.create(paper_dir, showWarnings = FALSE, recursive = TRUE)

DPI <- 1200

# Consistent condition color palette (used across all plots)
pal_control   <- "#4575B4"
pal_treatment <- "#D73027"
cond_palette  <- c("Control" = pal_control, "Treatment" = pal_treatment)
pal_diverging <- colorRampPalette(c("#4575B4", "white", "#D73027"))(100)
```

---

## 1. Load required intermediate data

```{r load_data, message = FALSE, warning = FALSE}
# Full DESeq2 results table (all genes, script 3)
res <- read.csv("../results/deseq_results_Treatment_vs_Control.csv", row.names = 1)

# Strict DEG subset (script 3)
deg_table <- read_csv("../results/deg_Treatment_vs_Control.csv", show_col_types = FALSE) %>%
  rename(GeneID_full = ...1) %>%
  mutate(GeneID = str_extract(GeneID_full, "WP_[0-9]+\\.[0-9]+"))
all_deg_ids <- trimws(deg_table$GeneID)

# Full annotated results (all genes + eggNOG, script 4)
full_annotation <- read_csv("../results/advanced_analysis/full_annotated_results.csv",
                            show_col_types = FALSE)

# eggNOG annotation (KEGG dictionary source)
eggnog_data <- read.delim("../data/annotation.tsv")

# Metadata and normalized counts (needed for the virulence heatmap)
metadatos <- read.csv(file = "../data/metadata_all.csv", row.names = 1)
load(file = "../results/dds_norm.RData")
normcounts <- counts(dds_norm, normalized = TRUE)
colnames(normcounts) <- metadatos$Replicates

# Gene-name label map (long ID -> short name, e.g. pelA)
label_map <- full_annotation %>%
  filter(!is.na(Preferred_name), !Preferred_name %in% c("NA", "-", "")) %>%
  dplyr::select(GeneID, Preferred_name) %>%
  distinct(GeneID, .keep_all = TRUE)

short_wp <- function(x) str_extract(x, "WP_[0-9]+\\.[0-9]+")
make_labels <- function(long_ids) {
  wp <- short_wp(long_ids)
  labels <- ifelse(wp %in% label_map$GeneID,
                   label_map$Preferred_name[match(wp, label_map$GeneID)],
                   wp)
  labels[is.na(labels)] <- wp[is.na(labels)]
  labels
}

# Virulence factors table (script 5_AdvancedAnalysis)
virulence_df <- read_csv("../results/advanced_analysis/virulence_factors_table.csv",
                         show_col_types = FALSE)
virulence_genes <- virulence_df$GeneID
```

---

## 2. Figure 1 — Volcano Plot: All Genes

```{r fig1_volcano_all, fig.width=11, fig.height=8.5, warning=FALSE}
row_labels <- make_labels(rownames(res))
sig_mask   <- !is.na(res$padj) & res$padj < 0.01 & abs(res$log2FoldChange) > 1

# --- Mixed ranking: combines significance (padj) and magnitude (log2FC) ---
# Uses rank() instead of raw normalized values, since many genes share the
# same clamped padj (~0), which would otherwise distort a min-max score.
# Takes pmin(rank_padj, rank_lfc) instead of summing, so a gene qualifies
# if it stands out strongly on EITHER axis — this prevents the hundreds of
# padj-clamped genes from dominating the selection and crowding out genes
# that are extreme only in log2FC.
N_TOP_LABELS <- 20

rank_df <- res[sig_mask, ] %>%
  as.data.frame() %>%
  rownames_to_column("GeneID_full") %>%
  mutate(
    neglog10_padj = -log10(padj),
    abs_lfc       = abs(log2FoldChange),
    rank_padj     = rank(neglog10_padj, ties.method = "average"),
    rank_lfc      = rank(abs_lfc, ties.method = "average"),
    best_rank     = pmin(rank_padj, rank_lfc)
  ) %>%
  arrange(desc(best_rank)) %>%
  slice_head(n = N_TOP_LABELS)

# --- Extra rule: always label genes with a very large fold change ---
# Independent of the ranking above, any significant gene whose |log2FC|
# exceeds this threshold gets labeled too, so large-effect outliers are
# never dropped just for missing the top-N cutoff.
FC_LABEL_THRESHOLD <- 2

fc_extreme_ids <- res[sig_mask, ] %>%
  as.data.frame() %>%
  rownames_to_column("GeneID_full") %>%
  filter(abs(log2FoldChange) > FC_LABEL_THRESHOLD) %>%
  pull(GeneID_full)

# --- Build a position-matched label vector (avoids selectLab name-collision) ---
# selectLab in EnhancedVolcano matches by label TEXT, so if two different
# rows share the same short gene symbol (e.g. two entries both mapping to
# "cysG"), it labels every occurrence of that text — including
# non-significant duplicates elsewhere in the table. Instead, we blank out
# all labels except the exact selected rows, matched by GeneID (rowname).
top_ids <- union(rank_df$GeneID_full, fc_extreme_ids)
plot_labels <- ifelse(rownames(res) %in% top_ids, row_labels, "")

volc_all <- EnhancedVolcano(res,
               lab             = plot_labels,
               selectLab       = plot_labels[plot_labels != ""],
               x               = "log2FoldChange",
               y               = "padj",
               axisLabSize     = 12,
               pCutoff         = 0.01,
               FCcutoff        = 1,
               col             = c("#BDBDBD", "#FDAE61", "#74ADD1", "#1A9641"),
               colAlpha        = 0.75,
               pointSize       = 2.5,
               title           = "Volcano Plot",
               titleLabSize    = 14,
               subtitle        = "Treatment vs Control — All Genes",
               subtitleLabSize = 11,
               captionLabSize  = 9,
               labSize         = 3.2,
               labFace         = "bold",
               labCol          = "black",
               legendLabels    = c("NS", "|log2FC| > 1", "adj-p < 0.01", "DEG"),
               legendPosition  = "right",
               legendLabSize   = 11,
               legendIconSize  = 4,
               drawConnectors  = TRUE,
               widthConnectors = 0.4,
               colConnectors   = "grey50",
               border          = "full",
               borderWidth     = 0.6,
               gridlines.major = TRUE,
               gridlines.minor = FALSE)
volc_all
ggsave(filename = file.path(paper_dir, "Fig1_Volcano_AllGenes.tiff"),
       plot = volc_all, width = 11, height = 8.5, dpi = DPI, compression = "lzw")
```

---

## 3. Figure 2 — KEGG Pathway Enrichment (GSEA Ridgeplot)

We run a Gene Set Enrichment Analysis (GSEA) over the full ranked gene list (by log2FoldChange)
against KEGG pathways, then plot the result as a ridgeplot. The "Photosynthesis" pathway(s) are
excluded from the plot, since they are not biologically relevant for this comparison.

```{r fig2_kegg, fig.width=10, fig.height=8, warning=FALSE, message=FALSE}
term2gene_kegg <- eggnog_data %>%
  dplyr::select(KEGG_Pathway, X.query) %>%
  tidyr::separate_rows(KEGG_Pathway, sep = ",") %>%
  dplyr::filter(KEGG_Pathway != "" & KEGG_Pathway != "-" & !is.na(KEGG_Pathway)) %>%
  dplyr::filter(grepl("map", KEGG_Pathway)) %>%
  dplyr::mutate(query = trimws(X.query), KEGG_Pathway = trimws(KEGG_Pathway)) %>%
  dplyr::select(KEGG_Pathway, X.query) %>%
  as.data.frame()

# Official KEGG pathway names, used as TERM2NAME for GSEA
kegg_names <- keggList("pathway")
term2name_kegg <- data.frame(
  PathwayID   = sub("path:", "", names(kegg_names)),
  PathwayName = as.character(kegg_names)
)
term2name_kegg$PathwayName <- gsub(" - Reference pathway", "", term2name_kegg$PathwayName)

# Build the ranked gene list (by log2FoldChange) required for GSEA
gsea_ranked <- full_annotation %>%
  filter(!is.na(GeneID) & !is.na(log2FoldChange)) %>%
  group_by(GeneID) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

gsea_vector <- gsea_ranked$log2FoldChange
names(gsea_vector) <- gsea_ranked$GeneID
gsea_vector <- sort(gsea_vector, decreasing = TRUE)

set.seed(124)
gsea_kegg <- GSEA(geneList = gsea_vector,
                  TERM2GENE = term2gene_kegg,
                  TERM2NAME = term2name_kegg,
                  pvalueCutoff = 0.05,
                  pAdjustMethod = "BH",
                  minGSSize = 10,
                  maxGSSize = 500,
                  eps = 0,
                  seed = TRUE)   # reproducible permutations (uses set.seed above)

# Drop any "Photosynthesis" pathways (case-insensitive) before plotting
gsea_kegg_filtered <- gsea_kegg
gsea_kegg_filtered@result <- gsea_kegg_filtered@result %>%
  dplyr::filter(!str_detect(tolower(Description), "photosynthesis"))

p_kegg_ridge <- ridgeplot(gsea_kegg_filtered, showCategory = 15) +
  labs(title = "KEGG Pathway Enrichment (GSEA)",
       x = "Enrichment Distribution (log2FoldChange)") +
  theme_minimal(base_size = 12) +
  theme(plot.title   = element_text(face = "bold", size = 14),
        axis.text.y  = element_text(size = 10),
        axis.text.x  = element_text(size = 10),
        legend.title = element_text(size = 10))
p_kegg_ridge

ggsave(filename = file.path(paper_dir, "Fig2_KEGG_Enrichment_Ridgeplot.tiff"),
       plot = p_kegg_ridge, width = 10, height = 8, dpi = DPI, compression = "lzw")
```

---

## 4. Figure 3 — Protein-Protein Interaction (PPI) Network

```{r fig3_ppi_build, warning=FALSE, message=FALSE}
deg_network <- full_annotation %>%
  filter(padj < 0.05) %>%
  dplyr::select(GeneID, Preferred_name, seed_ortholog, log2FoldChange) %>%
  mutate(search_name = coalesce(Preferred_name, GeneID)) %>%
  group_by(search_name) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

valid_orthologs <- deg_network$seed_ortholog[!is.na(deg_network$seed_ortholog) &
                                             deg_network$seed_ortholog != ""]
tax_id <- as.numeric(strsplit(valid_orthologs[1], "\\.")[[1]][1])

string_db <- STRINGdb$new(version = "11.5", species = tax_id, score_threshold = 700)
mapped_genes <- string_db$map(as.data.frame(deg_network), "search_name", removeUnmappedRows = TRUE)
interactions <- string_db$get_interactions(mapped_genes$STRING_id)

ppi_graph <- graph_from_data_frame(d = interactions, directed = FALSE)
ppi_graph <- simplify(ppi_graph)

if (vcount(ppi_graph) > 0) {
  # Louvain is stochastic: without a seed, module membership (and therefore the
  # per-module hub selection below) varies between runs. Seed 124 reproduces the
  # hub set shown in the published figure; 13 of the 15 hubs are stable across seeds.
  set.seed(124)
  communities <- cluster_louvain(ppi_graph)
  ppi_graph <- set_vertex_attr(ppi_graph, "cluster", value = as.numeric(membership(communities)))
  ppi_graph <- set_vertex_attr(ppi_graph, "degree", value = degree(ppi_graph))
  ppi_graph <- set_vertex_attr(ppi_graph, "betweenness", value = betweenness(ppi_graph))

  node_metadata <- data.frame(STRING_id = V(ppi_graph)$name) %>%
    left_join(mapped_genes, by = "STRING_id") %>%
    distinct(STRING_id, .keep_all = TRUE)

  ppi_graph <- set_vertex_attr(ppi_graph, "label", value = node_metadata$search_name)
  ppi_graph <- set_vertex_attr(ppi_graph, "log2FC", value = node_metadata$log2FoldChange)

  top_hubs <- node_metadata %>%
    mutate(Degree = V(ppi_graph)$degree,
           Betweenness = V(ppi_graph)$betweenness,
           Cluster = V(ppi_graph)$cluster) %>%
    group_by(Cluster) %>%
    slice_max(order_by = Betweenness, n = 3, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(desc(Betweenness)) %>%
    head(15)
} else {
  message("STRINGdb could not find any interactions above the 700 threshold.")
}
```

```{r fig3_ppi_plot, fig.width=12, fig.height=12, warning=FALSE}
if (vcount(ppi_graph) > 0) {
  net_components <- components(ppi_graph)
  if (max(net_components$csize) > 1) {
    main_component <- induced_subgraph(
      ppi_graph, vids = which(net_components$membership == which.max(net_components$csize)))

    tidy_net <- as_tbl_graph(main_component) %>%
      mutate(
        is_hub            = name %in% top_hubs$STRING_id,
        hub_label         = ifelse(is_hub, label, NA),
        clean_betweenness = replace_na(betweenness, 0),
        max_bet           = max(clean_betweenness, na.rm = TRUE),
        node_size         = ifelse(max_bet > 0, (clean_betweenness / max_bet) * 10 + 3, 5)
      )

    p_net <- ggraph(tidy_net, layout = "kk") +
      geom_edge_link(color = "grey85", width = 0.5, alpha = 0.6) +
      geom_node_point(aes(size = node_size, color = as.factor(cluster), alpha = is_hub)) +
      scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.4), guide = "none") +
      geom_node_text(aes(label = hub_label),
                     repel = TRUE, fontface = "bold", size = 5,
                     box.padding = 1, point.padding = 0.5, max.overlaps = Inf,
                     bg.color = "white", bg.r = 0.15,
                     segment.color = "grey40", segment.size = 0.6) +
      # --- Color legend: Louvain clusters are shown as visual grouping only,
      # not as biologically meaningful modules. Title reflects this so the
      # reader doesn't over-interpret the color as functional annotation. ---
      scale_color_viridis_d(option = "viridis", name = "Network module\n(visual grouping)",
                            labels = NULL, guide = guide_legend(
                              override.aes = list(size = 5),
                              label = FALSE,
                              keywidth = 0.6)) +
      # --- Size legend: betweenness centrality, shown as real continuous values ---
      scale_size_continuous(name = "Betweenness\ncentrality",
                            range = c(1.5, 11),
                            breaks = scales::breaks_pretty(n = 4)) +
      labs(title    = "Protein-Protein Interaction Network",
           subtitle = "Top 15 Bottleneck Hubs Highlighted (High Confidence STRING Score > 700)",
           # --- Clarifying caption: the analytical focus of this figure is
           # bottleneck identification (betweenness), not the cluster
           # colors, which only separate the network visually. ---
           caption  = "Node color denotes Louvain clusters used for visual separation only\nand does not imply shared biological function. Labeled nodes are the\ntop 15 bottleneck hubs by betweenness centrality.") +
      theme_void() +
      theme(legend.position = "right",
            legend.title    = element_text(face = "bold", size = 11),
            legend.text     = element_text(size = 9),
            legend.spacing  = unit(0.6, "cm"),
            plot.title      = element_text(face = "bold", size = 18, hjust = 0.5),
            plot.subtitle   = element_text(size = 14, hjust = 0.5, color = "grey40"),
            plot.caption    = element_text(size = 9, color = "grey40", hjust = 0.5,
                                           margin = margin(t = 10)),
            plot.margin     = margin(10, 10, 10, 10))

    # Render once as a high-res PDF, then rasterize to TIFF at 1200 dpi
    pdf_path  <- file.path(paper_dir, "Fig3_PPI_Network_tmp.pdf")
    tiff_path <- file.path(paper_dir, "Fig3_PPI_Network.tiff")

    pdf(pdf_path, width = 12, height = 12)
    print(p_net)
    dev.off()

    image_write(
      image_read_pdf(pdf_path, density = DPI),
      path = tiff_path, format = "tiff", compression = "lzw"
    )
    file.remove(pdf_path)

    print(p_net)
  } else {
    message("No connected components found to draw the plot.")
  }
} else {
  message("Plot cancelled: the network is empty.")
}
```

---

## 5. Figure 4 — Volcano Plot: Virulence Genes Highlighted

```{r fig4_virulence_volcano, fig.width=10, fig.height=8, warning=FALSE}
# Functional keywords are searched in the eggNOG description; the gene-family
# prefixes (pel, peh, out, kdg) are matched only at the start of the gene name,
# so that e.g. "out" does not capture every "outer membrane" protein.
virulence_desc_kw <- "pectate lyase|cellulase|protease|secretion system"
virulence_name_kw <- "^(pel|peh|out|kdg)"

# --- Relaxed significance cutoffs for virulence highlighting ---
# Original cutoffs (padj < 0.05, |log2FC| > 1 / 2-fold) were too strict and
# left almost all virulence-associated genes in the "NS" (not significant)
# bucket even when they showed meaningful shifts. Relaxed to the standard
# 1.5-fold change threshold (log2(1.5) ≈ 0.585) and a more permissive
# padj < 0.1, so more biologically relevant virulence genes are highlighted.
VIR_FC_CUTOFF   <- 0.585
VIR_PADJ_CUTOFF <- 0.1

virulence_volcano_data <- full_annotation %>%
  mutate(
    Is_Virulence = if_else(
      str_detect(tolower(Description), virulence_desc_kw) |
        str_detect(tolower(Preferred_name), virulence_name_kw), "Yes", "No"),
    Category = case_when(
      Is_Virulence == "Yes" & !is.na(padj) & padj < VIR_PADJ_CUTOFF & log2FoldChange >  VIR_FC_CUTOFF ~ "Virulence ↑ in Treatment",
      Is_Virulence == "Yes" & !is.na(padj) & padj < VIR_PADJ_CUTOFF & log2FoldChange < -VIR_FC_CUTOFF ~ "Virulence ↓ in Treatment",
      Is_Virulence == "Yes"                                                                            ~ "Virulence (NS)",
      TRUE                                                                                              ~ "Background"
    ),
    Label = if_else(
      Is_Virulence == "Yes" & !is.na(padj) & padj < VIR_PADJ_CUTOFF & abs(log2FoldChange) > VIR_FC_CUTOFF,
      coalesce(
        na_if(Preferred_name, "-"),
        str_extract(GeneID, "WP_[0-9]+\\.[0-9]+"),
        GeneID),
      NA_character_)
  )

vir_colors <- c("Background"                = "#BDBDBD",
                "Virulence (NS)"            = "#A6D96A",
                "Virulence ↑ in Treatment"   = "#D73027",
                "Virulence ↓ in Treatment" = "#4575B4")
vir_sizes  <- c("Background" = 1.2, "Virulence (NS)" = 2.5,
                "Virulence ↑ in Treatment" = 3.5, "Virulence ↓ in Treatment" = 3.5)
vir_alpha  <- c("Background" = 0.35, "Virulence (NS)" = 0.65,
                "Virulence ↑ in Treatment" = 1, "Virulence ↓ in Treatment" = 1)

p_vir_volcano <- ggplot(virulence_volcano_data, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = Category, size = Category, alpha = Category)) +
  scale_color_manual(values = vir_colors, name = NULL) +
  scale_size_manual( values = vir_sizes,  name = NULL) +
  scale_alpha_manual(values = vir_alpha,  name = NULL) +
  geom_text_repel(data = ~ filter(.x, !is.na(Label)),
                  aes(label = Label),
                  size = 3.5, fontface = "bold",
                  box.padding = 0.5, point.padding = 0.3,
                  max.overlaps = Inf,
                  color = "black", bg.color = "white", bg.r = 0.12,
                  segment.color = "grey50", segment.size = 0.4) +
  geom_vline(xintercept = c(-VIR_FC_CUTOFF, VIR_FC_CUTOFF), linetype = "dashed",
             color = "grey30", alpha = 0.6, linewidth = 0.5) +
  geom_hline(yintercept = -log10(VIR_PADJ_CUTOFF), linetype = "dashed",
             color = "grey30", alpha = 0.6, linewidth = 0.5) +
  labs(title    = "Virulence-Associated Genes",
       subtitle = "Red = higher in Treatment (log2FC > 0) · Blue = higher in Control (log2FC < 0)",
       x = "log₂ Fold Change  (Treatment / Control)",
       y = "-log₁₀(adj. p-value)") +
  theme_minimal(base_size = 13) +
  theme(plot.title       = element_text(face = "bold", size = 15),
        plot.subtitle    = element_text(size = 10, color = "grey40"),
        panel.grid.minor = element_blank(),
        legend.position  = "right")
p_vir_volcano
ggsave(filename = file.path(paper_dir, "Fig4_Volcano_VirulenceGenes.tiff"),
       plot = p_vir_volcano, width = 10, height = 8, dpi = DPI, compression = "lzw")
```

---

## 6. Figure 5 — Heatmap: Virulence Genes

```{r fig5_virulence_heatmap, fig.width=8, fig.height=10, warning=FALSE}
rownames(normcounts) <- str_extract(rownames(normcounts), "WP_[0-9]+\\.[0-9]+")
normcounts_clean <- normcounts[!is.na(rownames(normcounts)), ]

virulence_matrix <- normcounts_clean[virulence_genes, ]
rownames(virulence_matrix) <- ifelse(!is.na(virulence_df$Preferred_name),
                                     virulence_df$Preferred_name,
                                     virulence_df$GeneID)

p_virulence_hm <- pheatmap(virulence_matrix,
                           scale = "row",
                           color = viridis(100),
                           treeheight_row = 15,
                           treeheight_col = 15,
                           main = "Pectobacterium Virulence Factors Expression",
                           fontsize_row = 8)

ggsave(filename = file.path(paper_dir, "Fig5_Heatmap_VirulenceGenes.tiff"),
       plot = p_virulence_hm, width = 8, height = 10, dpi = DPI, compression = "lzw")
```

---

## 7. Figure 6 — Volcano Plot: Transcriptional Regulators

```{r fig6_regulators_volcano, fig.width=10, fig.height=8, warning=FALSE}
tf_keywords <- "transcription factor|regulator|dna-binding"   # matched against lower-cased text

regulator_volcano_data <- full_annotation %>%
  mutate(
    Is_Regulator = if_else(
      str_detect(tolower(Description), tf_keywords) |
        str_detect(tolower(Preferred_name), tf_keywords), "Yes", "No"),
    Category = case_when(
      Is_Regulator == "Yes" & !is.na(padj) & padj < 0.05 & log2FoldChange >  1 ~ "Regulator ↑ in Treatment",
      Is_Regulator == "Yes" & !is.na(padj) & padj < 0.05 & log2FoldChange < -1 ~ "Regulator ↓ in Treatment",
      Is_Regulator == "Yes"                                                      ~ "Regulator (NS)",
      TRUE                                                                       ~ "Background"
    ),
    Label = if_else(
      Is_Regulator == "Yes" & !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1,
      coalesce(
        na_if(Preferred_name, "-"),
        str_extract(GeneID, "WP_[0-9]+\\.[0-9]+"),
        GeneID),
      NA_character_)
  )

reg_colors <- c("Background"                = "#BDBDBD",
                "Regulator (NS)"            = "#9ECAE1",
                "Regulator ↑ in Treatment"   = "#D73027",
                "Regulator ↓ in Treatment" = "#4575B4")
reg_sizes  <- c("Background" = 1.2, "Regulator (NS)" = 2.5,
                "Regulator ↑ in Treatment" = 3.5, "Regulator ↓ in Treatment" = 3.5)
reg_alpha  <- c("Background" = 0.35, "Regulator (NS)" = 0.65,
                "Regulator ↑ in Treatment" = 1, "Regulator ↓ in Treatment" = 1)

p_tf_volcano <- ggplot(regulator_volcano_data, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = Category, size = Category, alpha = Category)) +
  scale_color_manual(values = reg_colors, name = NULL) +
  scale_size_manual( values = reg_sizes,  name = NULL) +
  scale_alpha_manual(values = reg_alpha,  name = NULL) +
  geom_text_repel(data = ~ filter(.x, !is.na(Label)),
                  aes(label = Label),
                  size = 3.5, fontface = "bold",
                  box.padding = 0.5, point.padding = 0.3,
                  max.overlaps = Inf,
                  color = "black", bg.color = "white", bg.r = 0.12,
                  segment.color = "grey50", segment.size = 0.4) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed",
             color = "grey30", alpha = 0.6, linewidth = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",
             color = "grey30", alpha = 0.6, linewidth = 0.5) +
  labs(title    = "Global Transcriptional Regulators",
       subtitle = "Red = higher in Treatment (log2FC > 0) · Blue = higher in Control (log2FC < 0)",
       x = "log₂ Fold Change  (Treatment / Control)",
       y = "-log₁₀(adj. p-value)") +
  theme_minimal(base_size = 13) +
  theme(plot.title       = element_text(face = "bold", size = 15),
        plot.subtitle    = element_text(size = 10, color = "grey40"),
        panel.grid.minor = element_blank(),
        legend.position  = "right")
p_tf_volcano

ggsave(filename = file.path(paper_dir, "Fig6_Volcano_Regulators.tiff"),
       plot = p_tf_volcano, width = 10, height = 8, dpi = DPI, compression = "lzw")
```

---

## Session information
```{r session_info}
sessionInfo()
```
